### 2021-11-18 manuscript/hiv-gwas.tex

\begin{flushleft}
	Sarah Nadeau\textsuperscript{1,2}, Christian W. Thorball\textsuperscript{3}, Roger Kouyos\textsuperscript{4,5}, Huldrych F. Günthard\textsuperscript{4,5}, Jürg Böni\textsuperscript{4}, Sabine Yerly\textsuperscript{6}, Matthieu Perreau\textsuperscript{7}, Thomas Klimkait\textsuperscript{8}, Andri Rauch\textsuperscript{9}, Hans H. Hirsch\textsuperscript{8,10,11}, Matthias Cavassini\textsuperscript{12}, Pietro Vernazza\textsuperscript{13}, Enos Bernasconi\textsuperscript{14}, Jacques Fellay\textsuperscript{2,3,15}, Venelin Mitov\textsuperscript{1,2,*}, Tanja Stadler\textsuperscript{1,2,*}, and the Swiss HIV Cohort Study (SHCS)
\end{flushleft}

\noindent
\textsuperscript{1}\emph{Department of Biosystems Science and Engineering, ETH Zürich, Basel} \\
\textsuperscript{2}\emph{Swiss Institute of Bioinformatics, Lausanne} \\
\textsuperscript{3}\emph{Precision Medicine Unit, Lausanne University Hospital and University of Lausanne, Lausanne, Switzerland} \\
\textsuperscript{4}\emph{Institute of Medical Virology, University of Zürich, Zürich} \\
\textsuperscript{5}\emph{Division of Infectious Diseases and Hospital Epidemiology, University Hospital Zürich, University of Zürich, Zürich} \\
\textsuperscript{6}\emph{Division of Infectious Diseases, Laboratory of Virology, Geneva University Hospital, Geneva} \\
\textsuperscript{7}\emph{Division of Immunology and Allergy, University Hospital Lausanne, Lausanne} \\
\textsuperscript{8}\emph{Department of Biomedicine, University of Basel, Basel} \\
\textsuperscript{9}\emph{Department of Infectious Diseases, Berne University Hospital and University of Berne, Berne} \\
\textsuperscript{10}\emph{Division of Clinical Virology, University Hospital Basel, Basel} \\
\textsuperscript{11}\emph{Division of Infectious Diseases and Hospital Epidemiology, University Hospital Basel, Basel} \\
\textsuperscript{12}\emph{Division of Infectious Diseases, University Hospital Lausanne, Lausanne} \\
\textsuperscript{13}\emph{Division of Infectious Diseases, Cantonal Hospital St. Gallen, St. Gallen} \\
\textsuperscript{14}\emph{Division of Infectious Diseases, Regional Hospital Lugano, Lugano} \\
\textsuperscript{15}\emph{Global Health Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne} \\
\textsuperscript{*}\emph{Co-last authors}

@@

\begin{abstract}
	Infectious diseases are a unique challenge for genome-wide association studies (GWAS) because pathogen, host, and environmental factors can all effect disease traits. Previous GWAS have successfully identified several human genetic variants associated with HIV-1 set point viral load (spVL), among other important infectious disease traits. However, these GWAS do not account for potentially confounding or extraneous pathogen effects that are heritable from donor to recipient in transmission chains.  We propose a novel method to consider the full genome of each patient's infecting pathogen strain, remove strain-specific effects on a trait based on the pathogen phylogeny, and thus better estimate the effect of human genetic variants on infectious disease traits. In simulations, we show our method can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations. When we apply our method to HIV-1 cohort data from Switzerland, we recover slightly weaker but qualitatively similar signals of association between spVL and human genetic variants in the \emph{CCR5} and major histocompatability complex (MHC) gene regions compared to standard GWAS. Our simulation study confirms that based on the estimated heritability and selection parameters for HIV-1 spVL, standard GWAS are robust to pathogen effects. Our framework may improve GWAS for other diseases where pathogen effects are stronger and more phylogenetically correlated amongst individuals in a cohort.
\end{abstract}

\section{Introduction}

@@

% Why is spVL and important trait
For HIV, GWAS have used a trait called set point viral load (spVL) to identify human variants associated with severity of disease course. spVL is generally defined to be the average concentration of viral particles in host plasma during the asymptomatic phase of infection in the absence of treatment \citep{Mellors1996}. In untreated individuals, spVL is predictive of duration of asymptomatic infection \citep{Mellors1996} and infectiousness \citep{Quinn2000}. If viral load can be reduced to undetectable levels, an individual is effectively uninfectious \citep{HHSARTGuidelines2019}. Notably, spVL varies by orders of magnitude between individuals \citep{Mellors1996}. Thus, spVL measurements point to a wide range in natural HIV control amongst individuals.

@@

% How heritable is spVL (human side)
Several GWAS have been done to measure the narrow-sense heritability of spVL and identify associated host genetic variants \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}. The largest study to-date by \citet{McLaren2015} estimated the narrow-sense heritability of spVL from human genetic variants to be approximately 25\%. All but 5\% of this was attributed to two regions in the human genome, the major histocompatibility complex (MHC) and C-C motif chemokine receptor 5 (\emph{CCR5}). Both associations are biologically relevant: the MHC encodes proteins that present viral epitopes at the cell surface and \emph{CCR5} encodes a co-receptor for HIV-1 cell entry. In other words, MHC proteins match bits of the virus like puzzle pieces and display these to signal that infected cells should be destroyed. CCR5 proteins help the virus infect target cells.

@@

In the following we (i) describe a statistical model for the spVL trait, (ii) derive a maximum likelihood estimate for the viral part of spVL under this model, and (iii) describe a new infectious disease GWAS framework using this information. In simulations, we demonstrate that this framework can improve GWAS power to detect host genetic variants effecting a disease trait. Finally, we apply our framework to human and viral genome data from the Swiss HIV Cohort Study (SHCS) and show that associations with spVL are robust to a correction for viral effects. Although we developed our framework in the context of HIV-1 spVL, the same approach could readily be applied to other heritable infectious disease traits.

\subsection{A statistical model for spVL}

Variation in spVL comes from several sources: direct host genetic effects, pathogen effects, interaction effects between the host and the pathogen, and other environmental effects. Of these, only pathogen effects are heritable from one transmission partner to another \citep{Leventhal2016}. To characterize these effects, we use a phylogenetic mixed model (PMM) \citep{Housworth2004}. PMMs assume continuous traits like spVL are the sum of independent heritable and non-heritable parts. In our case, pathogen effects comprise the heritable part and all other effects comprise the non-heritable part. Then, the heritable part is modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny. The non-heritable part is modeled as Gaussian noise added to sampled individuals at the tips of the phylogeny (Figure \ref{fig:spVLModel}).

\begin{figure}[H]
	\begin{center}
	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
% 		\includegraphics[width=0.462\linewidth]{figures/model_figure}
% 		\caption{The POUMM statistical model for the spVL trait. (A) shows how the pathogen effect on spVL, $g_v$, evolves according to an OU process along (C) the pathogen phylogeny. The dashed line in (A) shows the expected value of the trait at each point in time under the model. (B) shows that the non-pathogen effect on spVL, $e$, is modeled as a Gaussian random variable. Each tip on the phylogeny corresponds to a sampled individual with an independent non-pathogen-effect drawn from the distribution shown in (B).}
		\label{fig:spVLModel}
	\end{center}
\end{figure}

@@

\subsection{Teasing apart pathogen and non-pathogen effects on spVL}

@@

\subsection{A POUMM-based GWAS framework for infectious disease}

@@

\section{Results}

\subsection{Simulation study}

To test the theoretical best-case performance of our method, we simulated data under the POUMM and applied our framework to the simulated data. Figure \ref{fig:sim-design} shows the simulation scheme and table \ref{tab:sim-params} gives the value or expression for each parameter. In a nutshell, we simulated independent, additive host genetic effects, independent environmental effects, and heritable pathogen genetic effects under different scenarios of trait heritability and selection strength. To maintain the same heritability while varying selection strength, we counter-balanced by varying the intensity of stochastic evolutionary fluctuations accordingly. We fixed other variables to plausible values based on the spVL literature.

\subsubsection{Estimator accuracy}

First, we evaluated how well our method estimated the additive host genetic effects from the simulated data. Additive host genetic effects represent an ideal (albeit unattainable) baseline for infectious disease GWAS. Figure \ref{fig:simulationResults}A shows that our method incorporating phylogenetic information can more accurately estimate this value compared to the trait value. To ensure a fair comparison, we scaled trait values to have the same mean, zero, as host genetic effects so as not to bias the root mean squared error (RMSE) by a constant factor. In the supplemental material, we show why the scaled trait value is expected to have an RMSE of approximately 0.74 under our simulation set-up. By incorporating phylogenetic information, we can improve upon this error in scenarios where the trait is highly heritable and under low selection pressure. This is because high heritability means the pathogen greatly effects the trait value. Then, when the trait is under weak selection, pathogen effects can drift far from the long-term optimum and stochastic fluctuations are low to maintain the same heritability. Thus, our method performs well when an infectious disease trait is highly heritable and trait values are highly correlated amongst closely related members of a transmission chain, but vary widely within the entire cohort.

\subsubsection{Theoretical GWAS improvement}

@@

\subsection{GWAS on the Swiss HIV Cohort}

@@

% TODO
\begin{figure}[H]
	\begin{center}
		\includegraphics[width=0.7\linewidth]{figures/gwas_results}
		\caption{Results from comparative GWAS. (A) shows association p-values for the same host variants from the SHCS cohort in GWAS with two different response variables. On the left, we used unmodified spVL values. On the right, we used our estimates for the non-pathogen effects on spVL. The alternating shades correspond to different chromosomes. (B) compares the strength of association for variants in the \emph{CCR5} and MHC regions between the two GWAS (positions 45.4 - 47Mb on chromosome 3 and 29.5 - 33.5Mb on chromosome 6 for the \emph{CCR5} and MHC, respectively). Base positions are with reference to genome build GRCh37. The color of each point represents the difference in -log$_{10}$ p-value between the two GWAS. Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it. The dashed lines show genome-wide significance at p = 5x10$^{-8}$.}
		\label{fig:gwas-results}
	\end{center}
\end{figure}

@@

\section{Discussion}
% State the importance of your findings
In this paper, we presented a novel, phylogeny-aware GWAS framework to correct for heritable pathogen effects on infectious disease traits. By using information from the pathogen phylogeny, we show it is possible to improve GWAS power to detect host genetic variants associated with a disease trait. This should help us better understand which host factors are protective against a disease versus which increase susceptibility or disease severity.

% Contextualize the study & address limitations
Our method relies on the POUMM, a model of continuous trait evolution that accounts for heritable and non-heritable effects on a trait, as well as selection. Using this model, we estimated HIV-1 spVL heritability to be 45\% (95\% HPD 24 - 67\%) in the Swiss HIV Cohort Study. Compared to previous studies, this estimate is at the higher end \citep{Mitov2018}. Also using the POUMM, \citep{Bertels2018} estimated a spVL heritability of 29\% (N = 2014, CI 12 - 46\%) from the same cohort and \citep{Blanquart2017} estimated 31\% (N = 2028, CI 15 - 43\%) from a pan-European cohort. We note that our sample size (N = 1493 individuals) is quite a bit smaller than these other studies. This might be because we restricted samples based on having $pol$ gene sequences with at least 750 non-ambiguous bases. Our aim was to reconstruct a high-quality phylogeny, since the POUMM does not account for phylogenetic uncertainty and the POUMM parameter estimates are key to our downstream GWAS methods. Although our heritability estimate is rather high, the confidence interval largely overlaps that of other studies and we note that estimating heritability per se was not our primary focus.

@@

%\begin{itemize}
%	\item Compare inferred POUMM parameters to previous studies \citep{Mitov2018, Bertels2018}
%	\item Standard GWAS highlights previously identified regions associated with spVL: $CCR5$ and MHC
%	\item We expect non-viral only GWAS to increase signals of association in $CCR5$ (because noise in spVL from the virus reduced) and decrease signals of association in MHC (because viral mutations mediate MHC effects). It \emph{seems} like the results support these hypotheses, but I am waiting on POUMM parameter estimates from a longer MCMC run.
%	\item The phylogenetic correction to GWAS leads to the biggest p-value changes in the $CCR5$ and MHC regions
%	\item SNPs near the FUT1 gene (on chromosome 6, highlighted as one of the top-20 p-value changes in the first pipeline run) increased in significance, and the Lewis-X antigen produced by this gene is known to play a role in early HIV infection \citep{Naarding2005}.
%	\item Address method caveats: 
%	\begin{itemize}
%		\item We don't model virus-host interaction effects
%		\item We don't account for uncertainty in non-viral component of spVL estimates
%	\end{itemize}
%	\item Implications for future research: what other infectious disease traits are heritable?
%\end{itemize}

\section{Methods}

\subsection{Simulation model}

Whenever possible, we tried to parameterize our simulation model for spVL using empirical data. We set the total variance in spVL to 0.73 log copies$^2$ mL$^{-2}$ based on UK cohort data \citep{Mitov2018}. After allotting 25\% of this variance to the host part of spVL $g_h$ (following \citet{McLaren2015}), we partitioned the remaining variance between the viral part $g_v$ and the environmental part $\epsilon$ in different ratios to assess estimator performance across a range of spVL heritabilities. $g_h$ was simulated as the sum of contributions from 20 causal host genetic variants, 10 of which had an effect size of 0.2 log copies mL$^{-1}$ and 10 of which had an effect size of -0.2 log copies mL$^{-1}$. Host genetic variants were generated from a binomial distribution with probability $p$ calculated such that $g_h$ had the appropriate variance (see table \ref{tab:sim-params}). We generated a random viral phylogeny with branch lengths on the same time scale as a previously inferred UK cohort HIV tree \citep{Hodcroft2014} using the R package ape \citep{Paradis2018}. $g_v$ was simulated by running an OU process along the phylogeny using the R package POUMM \citep{Mitov2017a-POUMM} and sampling values at the tips. For the OU parameters $\theta$ and $g_{0}$ we used 4.5 log copies mL$^{-1}$, similar to values previously inferred for HIV (table \ref{tab:POUMM-parameters-our-estimates}). To assess our estimator's performance under a range of evolutionary scenarios, we co-varied the OU parameters for selection strength, $\alpha$, and intensity of random fluctuations, $\sigma$, so that different proportions of the variability in $g_v$ were attributable to selection and drift, respectively. Finally, the environmental component of spVL $\epsilon$ was generated from a normal distribution with mean 0.

\begin{table}[H]
	\begin{tabular}{p{1.5cm}ll} \toprule 
		Variable & Expression & Definition \\ \midrule 
		$\sigma^2_z$ &  0.73 log copies$^{2}$/mL$^{2}$ & total spVL variance \\ 
		$H^2_{\bar{t}}$ & varied & spVL viral heritability at $\bar{t}$ \\
		$\sigma^2_{g_h}$ & $\sigma_{g_h}^2 = 0.25*\sigma_z^2$ & variance in host part of spVL \\
		$\sigma^2_{g_v}(\bar{t})$ & $\sigma^2_{g_{v}}(\bar{t}) = H^2_{\bar{t}}*\sigma_z^2$ & variance in viral part of spVL at $\bar{t}$ \\ 
		$\sigma^2_{\epsilon}$ & $\sigma_{\epsilon}^2 = \sigma_z^2 - \sigma^2_{g_v} - \sigma^2_{g_h}$ & variance in environmental part of spVL \\ 
		$\bar{t}$ & 0.14 substitutions site$^{-1}$ yr$^{-1}$ & mean root-tip time in viral phylogeny \\ 
		$\bm{g}_v$ & $\bm{g}_v \sim Norm(\bm{\mu}_{OU}, \boldsymbol{\Sigma}_{OU})$ & viral part of spVL for all individuals \\ 
		$\theta$ & 4.5 log copies/mL & optimal spVL value \\
		$g_{0}$ & 4.5 log copies/mL & $g_v$ at the root of the phylogeny  \\ 
		$\alpha$ & varied & selection strength of OU process \\ 
		$\sigma$ & $\sigma = \sqrt{\frac{2\alpha\sigma^2_{g_{v}}(\bar{t})}{1 - exp(-2\alpha\bar{t})}}$ & time-unit standard deviation of OU process \\ 
		$\Psi$ & branch lengths $\sim Exp(\bar{t})$ & viral phylogeny \\ 
		$g_{h_i}$ & $g_{h_i} = \delta \sum_{j = 1}^{j = M/2}{G_{ij}} - \delta \sum_{j = M/2}^{j = M}{G_{ij}}$ & host part of spVL for individual $i$ \\ 
		$G_{N \times M}$ & $G_{ij} \sim Binom(2, p)\ \forall i \in {1...N}, \forall j \in {1...M}$ & host genotype matrix \\
		$p$ & $p = \frac{1}{2} - \sqrt{\frac{1}{4} - \frac{H_{\bar{t}}^2\sigma^2_z}{2\delta^2M}}$ &  host variant allele frequency \\ 
		$\delta$ & 0.2 & host variant effect size \\ 
		$M$ & 20 & number of causal host variants \\ 
		$\epsilon_i$ & $ \epsilon_i \sim Norm(0, \sigma^2_{\epsilon})$ & environmental part of spVL for individual $i$ \\ \bottomrule 
	\end{tabular}
	\caption{Simulation model parameters.}
	\label{tab:sim-params}
\end{table}

\subsection{Swiss HIV-1 data}

@@

\subsection{POUMM parameter inference}

@@

\subsection{Phylogenetic spVL correction}
We corrected calcualted spVL values using the method described in this paper. For each of the 1,392 individuals in the GWAS cohort, we estimated the viral part of spVL using equation \ref{eq:MuG} and the corresponding non-viral part using equation \ref{eq:EHat}. For the POUMM parameters $\alpha$, $\sigma$, $\theta$, and $\sigma_e$, we used the posterior mean estimates generated as described above.

>>>>

\begin{flushleft}
	Sarah Nadeau,$^{1,2}$ Christian W. Thorball,$^{3}$ Roger Kouyos,$^{4,5}$ Huldrych F. Günthard,$^{4,5}$ Jürg Böni,$^{4}$ Sabine Yerly,$^6$ Matthieu Perreau,$^{7}$ Thomas Klimkait,$^{8}$ Andri Rauch,$^{9}$ Hans H. Hirsch,$^{8,10,11}$ Matthias Cavassini,$^{12}$ Pietro Vernazza,$^{13}$ Enos Bernasconi,$^{14}$ Jacques Fellay,$^{2,3,15}$ Venelin Mitov,$^{\dagger,1,2}$ Tanja Stadler,$^{\dagger,\ast,1,2}$ and the Swiss HIV Cohort Study (SHCS)
\end{flushleft}

\noindent
$^{1}$Department of Biosystems Science and Engineering, ETH Zürich, Basel, Switzerland\\
$^{2}$Swiss Institute of Bioinformatics, Lausanne, Switzerland\\ 
$^{3}$Precision Medicine Unit, Lausanne University Hospital and University of Lausanne, Lausanne, Switzerland\\ 
$^{4}$Institute of Medical Virology, University of Zurich, Zurich, Switzerland\\ 
$^{5}$Division of Infectious Diseases and Hospital Epidemiology, University Hospital Zurich, University of Zurich, Zurich, Switzerland\\
$^{6}$Division of Infectious Diseases, Laboratory of Virology, Geneva University Hospital, Geneva, Switzerland\\
$^{7}$Division of Immunology and Allergy, University Hospital Lausanne, Lausanne, Switzerland\\
$^{8}$Department of Biomedicine, University of Basel, Basel, Switzerland\\
$^{9}$Department of Infectious Diseases, Bern University Hospital and University of Bern, Bern, Switzerland\\
$^{10}$Division of Clinical Virology, University Hospital Basel, Basel, Switzerland\\
$^{11}$Division of Infectious Diseases and Hospital Epidemiology, University Hospital Basel, Basel, Switzerland\\
$^{12}$Division of Infectious Diseases, University Hospital Lausanne, Lausanne, Switzerland\\
$^{13}$Division of Infectious Diseases, Cantonal Hospital St. Gallen, St. Gallen, Switzerland\\
$^{14}$Division of Infectious Diseases, Regional Hospital Lugano, Lugano, Switzerland\\
$^{15}$Global Health Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne, Lausanne, Switzerland\\
$\dagger$Co-last authors\\
$*$Corresponding author: tanja.stadler@bsse.ethz.ch

@@

\begin{abstract}
	Infectious diseases are a unique challenge for genome-wide association studies (GWAS) because pathogen, host, and environmental factors can all effect disease traits. Previous GWAS have successfully identified several human genetic variants associated with HIV-1 set point viral load (spVL), among other important infectious disease traits. However, these GWAS do not account for potentially confounding or extraneous pathogen effects that are heritable from donor to recipient in transmission chains.  We propose a novel method to consider the full genome of each patient's infecting pathogen strain, remove strain-specific effects on a trait based on the pathogen phylogeny, and thus better estimate the effect of human genetic variants on infectious disease traits. In simulations, we show our method can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations. When we apply our method to HIV-1 cohort data from the Swiss HIV Cohort Study, we recover slightly weaker but qualitatively similar signals of association between spVL and human genetic variants in the \emph{CCR5} and major histocompatibility complex (MHC) gene regions compared to standard GWAS. Our simulation study confirms that based on the estimated heritability and selection parameters for HIV-1 spVL, standard GWAS are robust to pathogen effects. Our framework may improve GWAS for other diseases where pathogen effects are stronger and more phylogenetically correlated amongst individuals in a cohort.
\end{abstract}

\section*{Introduction}

@@

% Why is spVL and important trait
For HIV, GWAS have used a trait called set point viral load (spVL) to identify human variants associated with severity of disease course. spVL is generally defined to be the average concentration of viral particles in host plasma during the asymptomatic phase of infection in the absence of treatment \citep{Mellors1996}. In untreated individuals, spVL is predictive of duration of asymptomatic infection \citep{Mellors1996} and infectiousness \citep{Quinn2000}. If viral load can be reduced to undetectable levels, an individual is effectively uninfectious and the risk of disease progression is massively reduced \citep{HHSARTGuidelines2019}. Notably, spVL varies by orders of magnitude between individuals \citep{Mellors1996}. Thus, spVL measurements point to a wide range in natural HIV control amongst individuals.

@@

% How heritable is spVL (human side)
Several GWAS have been done to measure the narrow-sense heritability of spVL and identify associated host genetic variants \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}. The largest study to-date by \citet{McLaren2015} estimated the narrow-sense heritability of spVL from human genetic variants to be approximately 25\%. All but 5\% of this was attributed to two regions in the human genome, the major histocompatibility complex (MHC) and C-C motif chemokine receptor 5 (\emph{CCR5}). Both associations are biologically relevant: the MHC encodes proteins that present viral epitopes at the cell surface and \emph{CCR5} encodes a co-receptor for HIV-1 cell entry. In other words, MHC proteins match bits of the virus like puzzle pieces and display these to signal that a cell is infected. CCR5 proteins help the virus infect target cells.

@@

In the following we describe a statistical model for the spVL trait, derive a maximum likelihood estimate for the viral part of spVL under this model, and describe a new infectious disease GWAS framework using this information. In simulations, we demonstrate that this framework can improve GWAS power to detect host genetic variants effecting a disease trait. Finally, we apply our framework to human and viral genome data from the Swiss HIV Cohort Study (SHCS) and show that associations with spVL are robust to a correction for viral effects. Although we developed our framework in the context of HIV-1 spVL, the same approach could readily be applied to other heritable infectious disease traits.

\section*{New Approaches}
\subsection*{A statistical model for spVL}

Variation in spVL comes from several sources: direct host genetic effects, pathogen effects, interaction effects between the host and the pathogen, and other environmental effects. Of these, only pathogen effects are heritable from one transmission partner to another \citep{Leventhal2016}. To characterize these effects, we use a phylogenetic mixed model (PMM) \citep{Housworth2004}. PMMs assume continuous traits like spVL are the sum of independent heritable and non-heritable parts. In our case, pathogen effects comprise the heritable part and all other effects comprise the non-heritable part. Then, the heritable part is modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny, as in Figure \ref{fig:spVLModel}A. The non-heritable part is modeled as Gaussian noise added to sampled individuals at the tips of the phylogeny.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
	\caption{A high-level schematic of our POUMM-based simulation framework. (A) shows how pathogen genetic effects on spVL evolve along the pathogen phylogeny according to an OU process. (B) shows host genetic effects, which are the sum of independent effects from several causal variants. Each variant can be present in 0, 1, or 2 copies and has some effect $\delta$ on the trait value. We assume each variant has the same magnitude of effect and that half the causal variants have a positive effect and the other half have a negative effect. (C) shows that environmental effects are assumed to be independent between sampled individuals and are drawn from a distribution centered at zero with variance $\sigma^2_e$.}
% 		\includegraphics[width=0.462\linewidth]{figures/model_figure}
% 		\caption{The POUMM statistical model for the spVL trait. (A) shows how the pathogen effect on spVL, $g_v$, evolves according to an OU process along (C) the pathogen phylogeny. The dashed line in (A) shows the expected value of the trait at each point in time under the model. (B) shows that the non-pathogen effect on spVL, $e$, is modeled as a Gaussian random variable. Each tip on the phylogeny corresponds to a sampled individual with an independent non-pathogen-effect drawn from the distribution shown in (B).}
		\label{fig:spVLModel}
	\end{center}
\end{figure}

@@

\subsection*{Teasing apart pathogen and non-pathogen effects on spVL}

@@

\subsection*{A POUMM-based GWAS framework for infectious disease}

@@

\section*{Results}

\subsection*{Simulation study}

To test the theoretical best-case performance of our method, we simulated data under the POUMM and applied our framework to the simulated data. Figure \ref{fig:spVLModel} shows a high-level schematic of our simulation framework and table \ref{tab:sim-params} gives the value or expression for each parameter. In a nutshell, we simulated independent, additive host genetic effects, independent environmental effects, and heritable pathogen genetic effects under different scenarios of trait heritability and selection strength. To maintain the same heritability while varying selection strength, we counter-balanced by varying the intensity of stochastic evolutionary fluctuations accordingly. We fixed other variables to plausible values based on the spVL literature.

\subsubsection*{Estimator accuracy}

First, we evaluated how well our method estimated the additive host genetic effects from the simulated data. Additive host genetic effects represent an ideal (albeit unattainable) baseline for infectious disease GWAS. Figure \ref{fig:simulationResults}A shows that our method incorporating phylogenetic information can more accurately estimate this value compared to the trait value. To ensure a fair comparison, we scaled trait values to have the same mean, zero, as host genetic effects so as not to bias the root mean squared error (RMSE) by a constant factor. In the supplemental material, we show why the scaled trait value is expected to have an RMSE of approximately 0.74 under our simulation scheme. By incorporating phylogenetic information, we can improve upon this error in scenarios where the trait is highly heritable and under low selection pressure. This is because high heritability means the pathogen greatly effects the trait value. Then, when the trait is under weak selection, pathogen effects can drift far from the long-term optimum and stochastic fluctuations are low to maintain the same heritability. Thus, our method performs well when an infectious disease trait is highly heritable and trait values are highly correlated amongst closely related members of a transmission chain, but vary widely within the entire cohort.

\subsubsection*{Theoretical GWAS improvement}

@@

\subsection*{GWAS on the Swiss HIV Cohort}

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=0.7\linewidth]{figures/gwas_results}
		\caption{Results from comparative GWAS. (A) shows association p-values for the same host variants from the SHCS cohort in GWAS with two different response variables. On the left, we used unmodified spVL values. On the right, we used our estimates for the non-pathogen effects on spVL. The alternating shades correspond to different chromosomes. (B) compares the strength of association for variants in the \emph{CCR5} and MHC regions between the two GWAS (positions 45.4 - 47Mb on chromosome 3 and 29.5 - 33.5Mb on chromosome 6 for the \emph{CCR5} and MHC, respectively). Base positions are with reference to genome build GRCh37. The color of each point represents the difference in -log$_{10}$ p-value between the two GWAS. Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it. The dashed lines show genome-wide significance at p = 5x10$^{-8}$.}
		\label{fig:gwas-results}
	\end{center}
\end{figure}

@@

\section*{Discussion}
% State the importance of your findings
In this paper, we presented a novel, phylogeny-aware GWAS framework to correct for heritable pathogen effects on infectious disease traits. By using information from the pathogen phylogeny, we show it is possible to improve GWAS power to detect host genetic variants associated with a disease trait. This should help us better understand which host factors are protective against a disease versus which increase susceptibility or disease severity.

% Contextualize the study & address limitations
Our method relies on the POUMM, a model of continuous trait evolution that accounts for heritable and non-heritable effects on a trait, as well as selection. Using this model, we estimated HIV-1 spVL heritability to be 45\% (95\% HPD 24 - 67\%) in the Swiss HIV Cohort Study. Compared to previous studies, this estimate is at the higher end \citep{Mitov2018}. Also using the POUMM, \citep{Bertels2018} estimated a spVL heritability of 29\% (N = 2014, CI 12 - 46\%) from the same cohort and \citep{Blanquart2017} estimated 31\% (N = 2028, CI 15 - 43\%) from a pan-European cohort. We note that our sample size (N = 1493 individuals) is smaller than in these other studies. This might be because we restricted samples based on having $pol$ gene sequences with at least 750 non-ambiguous bases. Our aim was to reconstruct a high-quality phylogeny, since the POUMM does not account for phylogenetic uncertainty and the POUMM parameter estimates are key to our downstream GWAS methods. Although our heritability estimate is rather high, the confidence interval largely overlaps that of other studies and we note that estimating heritability per se was not our primary focus.

@@

\section*{Materials and Methods}

\subsection*{Simulation model}

Whenever possible, we tried to parameterize our simulation model for spVL using empirical data. We set the total variance in spVL to 0.73 log copies$^2$ mL$^{-2}$ based on UK cohort data \citep{Mitov2018}. After allotting 25\% of this variance to the host part of spVL $g_h$ (following \citet{McLaren2015}), we partitioned the remaining variance between the viral part $g_v$ and the environmental part $\epsilon$ in different ratios to assess estimator performance across a range of spVL heritabilities. $g_h$ was simulated as the sum of contributions from 20 causal host genetic variants, 10 of which had an effect size of 0.2 log copies mL$^{-1}$ and 10 of which had an effect size of -0.2 log copies mL$^{-1}$. Host genetic variants were generated from a binomial distribution with probability $p$ calculated such that $g_h$ had the appropriate variance (see table \ref{tab:sim-params}). We generated a random viral phylogeny with branch lengths on the same time scale as a previously inferred UK cohort HIV tree \citep{Hodcroft2014} using the R package ape \citep{Paradis2018}. $g_v$ was simulated by running an OU process along the phylogeny using the R package POUMM \citep{Mitov2017a-POUMM} and sampling values at the tips. For the OU parameters $\theta$ and $g_{0}$ we used 4.5 log copies mL$^{-1}$, similar to values previously inferred for HIV (table \ref{tab:POUMM-parameters-our-estimates}). To assess our estimator's performance under a range of evolutionary scenarios, we co-varied the OU parameters for selection strength, $\alpha$, and intensity of random fluctuations, $\sigma$, so that different proportions of the variability in $g_v$ were attributable to selection and drift, respectively. Finally, the environmental component of spVL $\epsilon$ was generated from a normal distribution with mean 0.  For a full graphical model representation of the simulation scheme, see Figure \ref{fig:sim-design}.

\begin{table}[H]
	\begin{tabular}{p{1.5cm}ll} \toprule 
		Variable & Expression & Definition \\ \midrule 
		$\sigma^2_z$ &  0.73 log copies$^{2}$/mL$^{2}$ & Total spVL variance \\ 
		$H^2_h$ & 0.25 & Host heritability of spVL \\
		$H^2_{\bar{t}}$ & varied & Viral heritability of spVL at $\bar{t}$ \\
		$\sigma^2_{g_h}$ & $\sigma_{g_h}^2 = 0.25*\sigma_z^2$ & Variance in host part of spVL \\
		$\sigma^2_{g_v}(\bar{t})$ & $\sigma^2_{g_{v}}(\bar{t}) = H^2_{\bar{t}}*\sigma_z^2$ & Variance in viral part of spVL at $\bar{t}$ \\ 
		$\sigma^2_{\epsilon}$ & $\sigma_{\epsilon}^2 = \sigma_z^2 - \sigma^2_{g_v} - \sigma^2_{g_h}$ & Variance in environmental part of spVL \\ 
		$\bar{t}$ & 0.14 substitutions site$^{-1}$ yr$^{-1}$ & Mean root-tip time in viral phylogeny \\ 
		$\bm{g}_v$ & $\bm{g}_v \sim Norm(\bm{\mu}_{OU}, \boldsymbol{\Sigma}_{OU})$ & Viral part of spVL for all individuals \\ 
		$\theta$ & 4.5 log copies/mL & Optimal spVL value \\
		$g_{0}$ & 4.5 log copies/mL & $g_v$ at the root of the phylogeny  \\ 
		$\alpha$ & varied & Selection strength of OU process \\ 
		$\sigma$ & $\sigma = \sqrt{\frac{2\alpha\sigma^2_{g_{v}}(\bar{t})}{1 - exp(-2\alpha\bar{t})}}$ & Time-unit standard deviation of OU process \\ 
		$\Psi$ & branch lengths $\sim Exp(\bar{t})$ & Viral phylogeny \\ 
		$g_{h_i}$ & $g_{h_i} = \delta \sum_{j = 1}^{j = M/2}{G_{ij}} - \delta \sum_{j = M/2}^{j = M}{G_{ij}}$ & Host part of spVL for individual $i$ \\ 
		$G_{N \times M}$ & $G_{ij} \sim Binom(2, p)\ \forall i \in {1...N}, \forall j \in {1...M}$ & Host genotype matrix \\
		$p$ & $p = \frac{1}{2} - \sqrt{\frac{1}{4} - \frac{H_{h}^2\sigma^2_z}{2\delta^2M}}$ &  Host variant allele frequency \\ 
		$\delta$ & 0.2 & Host variant effect size \\ 
		$M$ & 20 & Number of causal host variants \\ 
		$\epsilon_i$ & $ \epsilon_i \sim Norm(0, \sigma^2_{\epsilon})$ & Environmental part of spVL for individual $i$ \\ 
		$N$ & 500 & Number of simulated samples \\ \bottomrule 
	\end{tabular}
	\caption{Simulation model parameters. For a full graphical model representation of the simulation scheme, including how these parameters are related, see Figure \ref{fig:sim-design}.}
	\label{tab:sim-params}
\end{table}

\subsection*{Swiss HIV-1 data}

@@

\subsection*{POUMM parameter inference}

@@

\subsection*{Phylogenetic spVL correction}
We corrected calcualted spVL values using the method described in this paper. For each of the 1,392 individuals in the GWAS cohort, we estimated the viral part of spVL using equation \ref{eq:MuG} and the corresponding non-viral part using equation \ref{eq:EHat}. For the POUMM parameters $\alpha$, $\sigma$, $\theta$, and $\sigma_e$, we used the posterior mean estimates generated as described above.
